Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments

### 2021-08-26 review/Response to reviews.Rmd

---
title: "Response to Reviews"
author: "Eric Scott"
date: "8/9/2021"
output: 
  bookdown::word_document2:
    reference_docx: template.docx
    number_sections: true
bibliography: references.bib
---

Cascading effects of increasing `k`:

-   Changed shape of crossbasis functions (edf changes, now not always more complex in fragments)

-   Changes in which lags are significant

-   Individual-level random effects impossible to fit due to computational limitations.
    Using a sub-sample, checked with and without invididual-level random effects.
    Changes r2, but very minimal changes to shape of smooths, edfs, and significance of smooths.
    Dropped from flowering model.

### 2021-08-26 doc/paper.Rmd
remove individual-level random effects (closes #96)

where $y_{ij}$ is the $i$th plant observation in the $j$th plot $s_1(z_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al. 2017 for details).
To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$ in @eq:eqn1.
We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$, in @eq:eqn1.
Additionally, in models for flowering probability and size, we included plant ID as a random effect to account for variation among individuals.
Preliminary analyses showed that this random effect was not significant in the growth models (edf \~ 0, p \> 0.2) and as such it was dropped to improve computational efficiency.
A random effect of plant ID was not included in survival models since each plant only dies once.
The crossbasis function, $s_2(\cdot)$ can also be written:

@@

*Flowering:* Overall, the the probability of flowering was significantly higher in continuous forest than in 1-ha fragments (Figure \@ref(fig:flwr-spei), Table \@ref(tab:intercept-table)).
SPEI had less of an effect on flowering probability in 1-ha fragments as indicated by the narrower range of the evaluated smooth (Figure \@ref(fig:flwr-spei)b).
This led to some important inter-habitat differences in plant responses to prior droughts.
In continuous forests, recent drought (i.e., at lag = 0--2 with SPEI \< -1), drought two dry seasons prior (lags 15--20) and in the wet season 34--36 months prior all increased the probability of flowering.
However, drought at the end of the rainy season one year prior (lags 7--13) significantly reduced flowering probability (Figure \@ref(fig:flwr-spei)a).
In fragments, recent drought had no significant effect on flowering probability and only drought at two dry seasons prior (lags 16--20) and in the wet season 33--36 months prior increased flowering probability (Figure \@ref(fig:flwr-spei)b).
The effects of drought on flowering probability were strong in continuous forest compared to 1 ha fragments (Figure \@ref(fig:flwr-spei)c).
We found no evidence for a cost of reproduction: in both forest and fragments, plants that had flowered in the previous year were on average more likely to be larger and flower again (Table \@ref(tab:results-table)).
The random effect of plant ID on flowering probability was also significant (Table \@ref(tab:results-table)), indicating significant individual-level variation in flowering probability.
<!--# individual level random effects need to be removed? -->

Finally, with the exception of the model for survival in 1-ha fragments, the random effect of plot was significant, indicating that vital rates varied among plots (Table \@ref(tab:results-table)).

@@

There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@eversLaggedDormantSeason2021], and hence be shaped by climatic events at any point in that physiological window.
For example, the flowering shoots of *Heliconia chartacea* begin to develop 6--10 months prior to the appearance of inflorescences [@crileyYearProductionHigh1994].
Adverse conditions during the 6 months following initiation, rather than the months when inflorescences are starting to expand, leads to the aborted production of flowering shoots.
Interestingly, we observed the opposite effect---drought conditions increased the probability of flowering two years later.
While this could reflect bet-hedging in response to stress [@nihadRelationshipStressFlowering2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988]. In fact, flowering in one year is associated with increased reproduction and growth in the next <!--# this may change once individual-level random effects are incorporated -->.

>>>>

where $y_{ij}$ is the $i$th plant observation in the $j$th plot $s_1(z_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al. 2017 for details).
To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$ in @eq:eqn1.
We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$, in @eq:eqn1.
A random effect of plant ID was not included in survival models since each plant only dies once.
Individual-level random effects were not included due to computational limitations in fitting random intercepts for all of the nearly 5,000 plants in continuous forest plots.
Preliminary modeling with data subsets showed that excluding individual-level random effects reduced model R^2^, but did not affect the shape or statistical significance of the smooths.

The crossbasis function, $s_2(\cdot)$ can also be written:

@@

*Flowering:* Overall, the the probability of flowering was significantly higher in continuous forest than in 1-ha fragments (Figure \@ref(fig:flwr-spei), Table \@ref(tab:intercept-table)).
SPEI had less of an effect on flowering probability in 1-ha fragments as indicated by the narrower range of the evaluated smooth (Figure \@ref(fig:flwr-spei)b).
This led to some important inter-habitat differences in plant responses to prior droughts.
In continuous forests, recent drought (i.e., at lag = 0--2 with SPEI \< -1), drought two dry seasons prior (lags 15--20) and in the wet season 34--36 months prior all increased the probability of flowering.
However, drought at the end of the rainy season one year prior (lags 7--13) significantly reduced flowering probability (Figure \@ref(fig:flwr-spei)a).
In fragments, recent drought had no significant effect on flowering probability and only drought at two dry seasons prior (lags 16--20) and in the wet season 33--36 months prior increased flowering probability (Figure \@ref(fig:flwr-spei)b).
The effects of drought on flowering probability were strong in continuous forest compared to 1 ha fragments (Figure \@ref(fig:flwr-spei)c).
We found no evidence for a cost of reproduction: in both forest and fragments, plants that had flowered in the previous year were on average more likely to be larger and flower again (Table \@ref(tab:results-table)).
Finally, with the exception of the model for survival in 1-ha fragments, the random effect of plot was significant, indicating that vital rates varied among plots (Table \@ref(tab:results-table)).

@@

There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@eversLaggedDormantSeason2021], and hence be shaped by climatic events at any point in that physiological window. For example, the flowering shoots of *Heliconia chartacea* begin to develop 6--10 months prior to the appearance of inflorescences [@crileyYearProductionHigh1994]. Adverse conditions during the 6 months following initiation, rather than the months when inflorescences are starting to expand, leads to the aborted production of flowering shoots. Interestingly, we observed the opposite effect---drought conditions increased the probability of flowering two years later. While this could reflect bet-hedging in response to stress [@nihadRelationshipStressFlowering2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988]. In fact, flowering in one year is associated with increased reproduction and growth in the next.
